---
output: github_document
---

# schemate <img src="man/figures/schemate.svg" align="right" />

> A small, [checkmate](https://mllg.github.io/checkmate/)-first schema DSL for R data.

[![CRAN_Status_Badge](https://www.r-pkg.org/badges/version/schemate)](http://cran.r-project.org/package=schemate)
[![CRAN RStudio mirror downloads](https://cranlogs.r-pkg.org/badges/schemate)](http://cran.r-project.org/web/packages/schemate/index.html)

`schemate` provides a small, [checkmate](https://mllg.github.io/checkmate/)-first
schema DSL for R data. It can infer schemas from example objects, edit schema
documents, save them as JSON, read them back, and validate new inputs against
the schema.

The package is meant for package authors and pipeline authors who want a compact
R-native schema format without adopting the full JSON Schema vocabulary. A
typical workflow is:

1. infer a conservative schema with `schema_infer()`;
2. edit it with `schema_*()` authoring verbs;
3. save it with `schema_write()`;
4. read it back with `schema_read()`;
5. validate inputs with `schema_validate()`.

## Installation

Currently, you can install the development version from GitHub.

```{r gh-install, eval = FALSE}
pak::pak("hongyuanjia/schemate")
```


## Quick Start

The public API uses a single `schema_` prefix and works well in pipelines.
Start from an example object, infer a conservative schema, then compact it into
something easier to edit and review.

```{r quick-start}
library(schemate)

payload <- list(
    items = list(
        list(id = 1L, name = "alpha", label = "Alpha", slug = "alpha"),
        list(id = 2L, name = "beta", label = "Beta", slug = "beta")
    )
)

schema <- payload |>
    schema_infer(keys = "named", arrays = "rest") |>
    schema_compact() |>
    schema_set_desc("$items", "Repository-like result items")

schema

schema |>
    schema_validate(payload, mode = "test")
```

`schema_validate()` defaults to assert mode: invalid input raises an error and
valid input is returned invisibly. Other modes are available when you need a
message or a boolean result.

```{r validation-modes}
bad_payload <- payload
bad_payload$items[[1L]]$id <- "bad"

schema |>
    schema_validate(bad_payload, mode = "check", name = "payload")
schema |>
    schema_validate(bad_payload, mode = "test", name = "payload")
```

For a data frame example, see the [Get started article](https://hongyuanjia.github.io/schemate/articles/schemate.html).

## JSON Workflow

Schemas are stored as a compact JSON DSL. The DSL is not JSON Schema; it is a
thin representation of checkmate checks, field schemas, local definitions, and
combinators. See the [Schema DSL article](https://hongyuanjia.github.io/schemate/articles/schema-dsl.html)
for the complete format reference. `schema_read()` and `schema_write()` require
the suggested package [jsonlite](https://github.com/jeroen/jsonlite).

```{r json-workflow}
path <- tempfile(fileext = ".json")
schema_write(schema, path)

restored <- schema_read(path)
restored

restored |>
    schema_validate(payload)
```

Example schema files are installed under `inst/extdata`:

```{r extdata, eval = FALSE}
system.file("extdata", "person-schema.json", package = "schemate")
```

## Validation Modes

`schema_validate()` supports four modes:

| Mode | Return value on success | Return value on failure |
| --- | --- | --- |
| `assert` | invisibly returns the input | throws an error |
| `check` | `TRUE` | diagnostic string |
| `test` | `TRUE` | `FALSE` |
| `expect` | testthat-style expectation object | expectation failure object |

Use `assert` inside application code, `check` when displaying diagnostics,
`test` for control flow, and `expect` in tests.

## Standalone Use

`schemate` also publishes a generated standalone bundle for packages that want
the schema features without depending on `schemate` at runtime.

```{r standalone, eval = FALSE}
usethis::use_standalone("hongyuanjia/schemate", "schema", ref = "standalone")
```

## Relation to Other Tools

`schemate` is closest in spirit to checkmate: schemas ultimately validate R
objects by calling checkmate checks. It adds a schema lifecycle around those
checks: infer, edit, serialize, read, and validate.

[`pointblank`](https://rstudio.github.io/pointblank/) is a better fit for
tabular data quality workflows, reporting, and column-oriented validation plans.
`schemate` is deliberately narrower and more structural: it describes R values,
R object names, nested lists, JSON-like payloads, and package-facing input
contracts. It is not a replacement for
[JSON Schema](https://json-schema.org/) or
[`jsonvalidate`](https://docs.ropensci.org/jsonvalidate/), which are better
choices when you need standards-compliant JSON document validation.

The R validation ecosystem is broad:

- [`validate`](https://cran.r-project.org/package=validate) captures data
  validation rules that can be documented, stored, and applied to data sets.
- [`assertr`](https://docs.ropensci.org/assertr/) is designed for assertive data
  checks inside analysis pipelines.
- [`data.validator`](https://appsilon.github.io/data.validator/) focuses on
  dataset validation with reporting.
- [`vetr`](https://cran.r-project.org/package=vetr) provides template-based
  structural checks for R objects.
- [`testthat`](https://testthat.r-lib.org/) is the right home for unit-test
  expectations; `schema_validate(..., mode = "expect")` is intended to fit into
  that style.

## License

The project is released under the terms of MIT License.
