---
title: "API Response Examples"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{API Response Examples}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

Real API responses often contain nested JSON arrays. A useful workflow is:

1. read the response into R with `jsonlite`;
2. infer an observed schema with `arrays = "rest"`;
3. compact repeated alternatives into a maintainable shape;
4. refine the few places where the API contract is stricter than the example.

Use `simplifyVector = FALSE` when the goal is to validate the original JSON-like
structure. If jsonlite simplifies arrays into vectors or data frames, that
simplified R object is still valid to check, but it is no longer a direct view of
the raw JSON response.

```{r setup, include = FALSE}
library(schemate)

payload_path <- function(file) {
    path <- system.file("extdata/api", file, package = "schemate")
    if (nzchar(path)) {
        return(path)
    }

    file.path("..", "inst", "extdata", "api", file)
}
```

## GitHub Search

For live data, read a GitHub REST API response without simplifying JSON arrays
into data frames:

```{r, eval = FALSE}
github_payload <- jsonlite::fromJSON(
    "https://api.github.com/search/repositories?q=language:R+schema",
    simplifyVector = FALSE,
    simplifyDataFrame = FALSE,
    simplifyMatrix = FALSE
)
```

This vignette uses a curated subset of a real saved response so it can run
without network access. The source URL and capture details are stored beside the
JSON file in `inst/extdata/api/SOURCE.md`.

```{r}
github_payload <- jsonlite::fromJSON(
    payload_path("github-search-repositories.json"),
    simplifyVector = FALSE,
    simplifyDataFrame = FALSE,
    simplifyMatrix = FALSE
)
jsonlite::toJSON(github_payload, pretty = TRUE, auto_unbox = TRUE)

github_schema <- github_payload |>
    schema_infer(keys = "required", arrays = "rest") |>
    schema_compact()
github_schema
```

The `items` field is inferred as an unnamed list whose `rest` schema describes
repository-like objects.

```{r}
bad_github <- github_payload
bad_github$items[[1L]]$owner$id <- "not an integer"

github_schema |>
    schema_validate(github_payload, mode = "test")
github_schema |>
    schema_validate(bad_github, mode = "check", name = "github")
```

## CrossRef Works

For live CrossRef data:

```{r, eval = FALSE}
crossref_payload <- jsonlite::fromJSON(
    "https://api.crossref.org/works?query.title=schema&rows=2",
    simplifyVector = FALSE,
    simplifyDataFrame = FALSE,
    simplifyMatrix = FALSE
)
```

Again, the runnable example uses a curated subset of a real saved response.

```{r}
crossref_payload <- jsonlite::fromJSON(
    payload_path("crossref-works.json"),
    simplifyVector = FALSE,
    simplifyDataFrame = FALSE,
    simplifyMatrix = FALSE
)
jsonlite::toJSON(crossref_payload, pretty = TRUE, auto_unbox = TRUE)

crossref_schema <- crossref_payload |>
    schema_infer(keys = "named", arrays = "rest") |>
    schema_compact()
crossref_schema
```

CrossRef `date-parts` is a tuple-like array. Inference records the observed
array shape, then a small manual edit can express year/month/day positions for
the saved response's `published-online` date.

The path below walks through the inferred homogeneous arrays. The first `rest`
means "the schema for each item"; the final `rest` means "the schema for each
date-parts entry".

```text
$message$items$rest$`published-online`$`date-parts`$rest
```

```{r}
date_part <- list(
    check = list(kind = "list", min.len = 1L, max.len = 3L),
    keys = list(type = "unnamed"),
    positions = list(
        schema_check("int", lower = 0),
        schema_check("int", lower = 1, upper = 12),
        schema_check("int", lower = 1, upper = 31)
    )
)

crossref_schema <- crossref_schema |>
    schema_replace("$message$items$rest$`published-online`$`date-parts`$rest", date_part)

bad_crossref <- crossref_payload
bad_crossref$message$items[[1L]]$`published-online`$`date-parts`[[1L]][[2L]] <- 13L

crossref_schema |>
    schema_validate(crossref_payload, mode = "test")
crossref_schema |>
    schema_validate(bad_crossref, mode = "check", name = "crossref")
```

When a reusable component has a clear domain name, author it explicitly with
`$defs` and `$ref`. `schema_compact()` deliberately avoids inventing reusable
definition names for you.
