---
title: "API Examples"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{API Examples}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

Real API responses often contain nested JSON arrays. A useful workflow is:

1. read the response into R with `jsonlite`;
2. infer an observed schema with `arrays = "rest"`;
3. compact repeated alternatives into a maintainable shape;
4. refine the few places where the API contract is stricter than the example.

```{r}
library(schemate)

payload_path <- function(file) {
  path <- system.file("extdata/api", file, package = "schemate")
  if (nzchar(path)) {
    return(path)
  }

  file.path("..", "inst", "extdata", "api", file)
}
```

## GitHub Search

For live data, read a GitHub REST API response without simplifying JSON arrays
into data frames:

```{r, eval = FALSE}
github_payload <- jsonlite::fromJSON(
  "https://api.github.com/search/repositories?q=language:R+schema",
  simplifyDataFrame = FALSE,
  simplifyMatrix = FALSE
)
```

This vignette uses a curated subset of a real saved response so it can run
without network access. The source URL and capture details are stored beside the
JSON file in `inst/extdata/api/SOURCE.md`.

```{r}
github_payload <- jsonlite::fromJSON(
  payload_path("github-search-repositories.json"),
  simplifyVector = FALSE,
  simplifyDataFrame = FALSE,
  simplifyMatrix = FALSE
)

github_schema <- schema_infer(github_payload, keys = "named", arrays = "rest")
github_schema <- schema_compact(github_schema)
as.list(github_schema)$fields$items
```

The `items` field is inferred as an unnamed list whose `rest` schema describes
repository-like objects.

```{r}
bad_github <- github_payload
bad_github$items[[1L]]$owner$id <- "not an integer"

schema_validate(github_schema, github_payload, mode = "test")
schema_validate(github_schema, bad_github, mode = "check", name = "github")
```

## CrossRef Works

For live CrossRef data:

```{r, eval = FALSE}
crossref_payload <- jsonlite::fromJSON(
  "https://api.crossref.org/works?query.title=schema&rows=2",
  simplifyDataFrame = FALSE,
  simplifyMatrix = FALSE
)
```

Again, the runnable example uses a curated subset of a real saved response.

```{r}
crossref_payload <- jsonlite::fromJSON(
  payload_path("crossref-works.json"),
  simplifyVector = FALSE,
  simplifyDataFrame = FALSE,
  simplifyMatrix = FALSE
)

crossref_schema <- schema_infer(crossref_payload, keys = "named", arrays = "rest")
crossref_schema <- schema_compact(crossref_schema)
as.list(crossref_schema)$fields$message$fields$items
```

CrossRef `date-parts` is a tuple-like array. Inference records the observed
array shape, then a small manual edit can express year/month/day positions for
the saved response's `created` date.

```{r}
date_part <- list(
  check = list(kind = "list", min.len = 1L, max.len = 3L),
  keys = list(type = "unnamed"),
  positions = list(
    schema_check("int", lower = 0),
    schema_check("int", lower = 1, upper = 12),
    schema_check("int", lower = 1, upper = 31)
  )
)

crossref_schema <- schema_replace(
  crossref_schema,
  "$message$items$rest$created$`date-parts`$rest",
  date_part
)

bad_crossref <- crossref_payload
bad_crossref$message$items[[1L]]$created$`date-parts`[[1L]][[2L]] <- 13L

schema_validate(crossref_schema, crossref_payload, mode = "test")
schema_validate(crossref_schema, bad_crossref, mode = "check", name = "crossref")
```

When a reusable component has a clear domain name, author it explicitly with
`$defs` and `$ref`. `schema_compact()` deliberately avoids inventing reusable
definition names for you.
