---
title: "Schema Workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Schema Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

`schemate` is built around one common workflow:

1. infer a conservative schema from an example object;
2. edit the schema until it expresses the real input contract;
3. write the schema to JSON;
4. read the schema back where it is needed;
5. validate new input.

## Infer

```{r}
library(schemate)

person <- list(id = 1L, name = "alice")
schema <- schema_infer(person, keys = "required")
schema
```

Inference captures structure and checkmate kinds. It does not guess business
rules, optional fields, reusable definitions, or descriptions.

## Edit

```{r}
schema <- schema_set_desc(schema, "$", "Person-like named list.")
schema <- schema_set_desc(schema, "$id", "Stable integer identifier.")
schema <- schema_set_desc(schema, "$name", "Display name.")
schema <- schema_add_field(
  schema,
  "email",
  schema_check("string", min.chars = 3, null.ok = TRUE),
  overwrite = TRUE
)
```

Paths use `$` for the root node. Bare field paths such as `$id` are shorthand for
the corresponding field under a container node.

## Write And Read

`schema_read()` and `schema_write()` require the suggested package `jsonlite`.

```{r}
path <- tempfile(fileext = ".json")
schema_write(schema, path)

restored <- schema_read(path)
restored
```

Package authors can store schema files in `inst/extdata` and load them with
`system.file()`.

```{r}
person_schema <- system.file("extdata", "person-schema.json", package = "schemate")
schema_read(person_schema)
```

## Validate

```{r}
good <- list(id = 2L, name = "bob", email = "bob@example.com")
schema_validate(restored, good)

bad <- list(id = "two", name = "bob")
schema_validate(restored, bad, mode = "check")
```

Use validation modes according to the caller:

| Mode | Use when |
| --- | --- |
| `assert` | invalid input should stop execution |
| `check` | you want a diagnostic string |
| `test` | control flow needs a boolean |
| `expect` | tests should receive an expectation object |

## Interpret Errors

Diagnostics include a path prefix. For example, a message starting with
`bad$id` means the root object named `bad` failed at its `id` field. Messages
from leaf checks come from checkmate, while container messages are produced by
`schemate` when fields, names, branches, or references do not match.
