---
title: "Schema DSL"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Schema DSL}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(schemate)
```

This vignette documents the JSON DSL used by `schemate`. Most users should start
with `schema_infer()` and the edit helpers; hand-written DSL is useful for
package fixtures, configuration files, and stable schemas reviewed outside R.

# schemate JSON DSL

## Goal

This DSL is a human-centric, checkmate-first schema format for hand-written JSON.
It is designed to describe the structure of an existing R object and to be parsed
into a `schemate` `SchemaDoc` object.

This DSL is **not** a JSON Schema implementation. Its semantics are derived from
`checkmate::assert_*()` and a small set of schema combinators.

## Minimum mental model

Most schema documents are nested combinations of a few keywords:

| DSL key | Use it for |
| --- | --- |
| `check` | the checkmate constraint for the current node |
| `fields` | exact field-name schemas for named containers |
| `groups` | multiple sibling fields that share the same schema |
| `patterns` | regex field-name schemas for otherwise unspecified fields |
| `positions` | position-specific schemas for unnamed lists |
| `rest` | otherwise unspecified fields or remaining unnamed-list positions |
| `$defs` / `$ref` | manually named reusable schema components |
| `all`, `any`, `one`, `not` | schema combinators |

For most workflows, start with `schema_infer()`, inspect `as.list(schema)`, then
use edit verbs to add the few rules inference cannot know.

## Validation examples

Most examples below show the JSON DSL shape first, then read the same JSON text
as a schema:

```r
schema <- schema_read('{ ... }')
schema_validate(schema, good, mode = "test", name = "payload")
schema_validate(schema, bad, mode = "check", name = "payload")
```

The `test` call shows a valid input as a boolean result. The `check` call shows
the diagnostic string for an invalid input without interrupting the vignette.

## Common schema shapes

A scalar schema is just a `check` node:

```json
{
  "check": { "kind": "string", "min.chars": 1 }
}
```

```{r}
schema <- schema_read('{
  "check": { "kind": "string", "min.chars": 1 }
}')
good <- "run-001"
bad <- ""

schema_validate(schema, good, mode = "test", name = "payload")
schema_validate(schema, bad, mode = "check", name = "payload")
```

A named container uses `fields` for exact child names and can use `rest` for the
remaining fields:

```json
{
  "check": { "kind": "list" },
  "keys": { "type": "named" },
  "fields": {
    "id": { "check": { "kind": "int", "lower": 1 } }
  },
  "rest": { "check": { "kind": "string" } }
}
```

```{r}
schema <- schema_read('{
  "check": { "kind": "list" },
  "keys": { "type": "named" },
  "fields": {
    "id": { "check": { "kind": "int", "lower": 1 } }
  },
  "rest": { "check": { "kind": "string" } }
}')
good <- list(id = 1L, label = "alpha")
bad <- list(id = 1L, label = 2L)

schema_validate(schema, good, mode = "test", name = "payload")
schema_validate(schema, bad, mode = "check", name = "payload")
```

An unnamed JSON-like array uses `keys.type = "unnamed"`. Use `rest` for a
homogeneous array and `positions` when the first positions have tuple-like
meaning:

```json
{
  "check": { "kind": "list" },
  "keys": { "type": "unnamed" },
  "rest": { "check": { "kind": "string" } }
}
```

```{r}
schema <- schema_read('{
  "check": { "kind": "list" },
  "keys": { "type": "unnamed" },
  "rest": { "check": { "kind": "string" } }
}')
good <- list("alpha", "beta")
bad <- list("alpha", 2L)

schema_validate(schema, good, mode = "test", name = "payload")
schema_validate(schema, bad, mode = "check", name = "payload")
```

The rest of this article is a reference for the same building blocks. The
container-child sections explain `fields`, `groups`, `patterns`, `positions`,
and `rest`; the formal-rule sections explain reserved keys, shorthand syntax,
validation order, and serialization order.

## When to write DSL JSON by hand

Most users should start from `schema_infer()`, edit the result with
`schema_add_field()`, `schema_set_desc()`, `schema_set_keys()`, and related
helpers, then save with `schema_write()`. Hand-written JSON is useful when:

- a schema is part of a package or analysis configuration;
- a non-R workflow needs to review or patch the schema text;
- you want stable schema fixtures for tests;
- you are translating an existing input contract into a compact R schema.

Prefer small, explicit schemas. Use `$defs` and `$ref` when the same child schema
appears repeatedly, and use `groups` when several sibling fields share one rule.

## Helper mapping

The R helpers create the same JSON shapes documented below:

| R helper | DSL keyword |
| --- | --- |
| `schema_check()` | `check` |
| `schema_ref()` | `$ref` |
| `schema_all()` | `all` |
| `schema_any()` | `any` |
| `schema_one()` | `one` |
| `schema_not()` | `not` |
| `schema_group()` | `groups[]` |

## Formal reference

### Core principles

1. Each schema node has exactly one **primary operator**.
2. The primary operator must be one of:
   - `check`
   - `all`
   - `any`
   - `one`
   - `not`
   - `$ref`
3. A `check` node uses `check.kind` to select a supported
   `checkmate::assert_*()` suffix.
4. All keys inside `check` other than `kind` are passed directly to
   `assert_<kind>()`.
5. There is no `checks` keyword in this grammar; use `all` for conjunction.
6. Container kinds are:
   - `list`
   - `data_frame`
   - `data_table`
   - `tibble`
7. `fields`, `groups`, `patterns`, `positions`, `rest`, and `keys` are only
   valid on `check` nodes whose `check.kind` is a container kind.
8. `positions` maps 1-based positions to child schemas for unnamed lists.
9. `rest` is the child schema for otherwise unspecified object fields or
   unnamed list elements.
10. `patterns` maps regular expressions to child schemas for otherwise
   unspecified object fields whose names match those patterns.
11. `groups` can batch-assign multiple field names to one shared child schema
   node.
12. A top-level schema document may optionally declare `version` as a
    non-empty string representing the document version.
13. Root-level `$defs` can store reusable schema nodes and `$ref` can reuse them.
14. Inside `all` / `any` / `one` / `not`, a branch may use shorthand check
    syntax without an explicit `check` wrapper.

### Reserved schema DSL keys

The following keys are reserved by the schema DSL:

- `check`
- `all`
- `any`
- `one`
- `not`
- `fields`
- `groups`
- `patterns`
- `positions`
- `rest`
- `keys`
- `description`
- `$defs`
- `$ref`
- `version`

`$defs` and `version` are document-level keys. They are only valid on the
top-level schema document and must not appear inside nested schema nodes or
shorthand combinator branches.

Field names and field-name rules from the validated data do **not** appear at
the node top level. They appear inside:

- `fields`
- `groups[].names`
- `patterns`
- `$defs` definition names

### Top-level schema document

A schema document contains:

- an optional `version` string
- an optional `$defs` object
- exactly one root schema node

Example:

```json
{
  "version": "1.0.0",
  "$defs": {
    "text": { "check": { "kind": "string" } }
  },
  "description": "root alias",
  "$ref": "#/$defs/text"
}
```

`version` is document metadata only. It describes the schema document version
and does not alter node parsing or validation semantics.

For canonical serialization via `as.list()`, top-level keys are emitted in this
order when present:

1. `version`
2. `$defs`
3. the root schema node keys

## Container check nodes

Only the following `check.kind` values are treated as container checks:

- `list`
- `data_frame`
- `data_table`
- `tibble`

Only these container checks may carry:

- `keys`
- `fields`
- `groups`
- `patterns`
- `positions`
- `rest`

All non-container `check.kind` values must reject these keys.

## Node forms

### Check node

A check node has the form:

```json
{
  "check": {
    "kind": "string",
    "pattern": "^[0-9]+$",
    "null.ok": true
  }
}
```

```{r}
schema <- schema_read('{
  "check": {
    "kind": "string",
    "pattern": "^[0-9]+$",
    "null.ok": true
  }
}')
good <- "123"
bad <- "abc"

schema_validate(schema, good, mode = "test", name = "payload")
schema_validate(schema, bad, mode = "check", name = "payload")
```

Rules:

- `check` must be an object.
- `check.kind` must be a non-empty string.
- `check.kind` must name a supported `checkmate::assert_*()` suffix.
- All other keys inside `check` are forwarded to `assert_<kind>()`.

Examples:

- `"string"` -> `assert_string()`
- `"list"` -> `assert_list()`
- `"choice"` -> `assert_choice()`
- `"int"` -> `assert_int()`
- `"number"` -> `assert_number()`
- `"flag"` -> `assert_flag()`
- `"data_frame"` -> `assert_data_frame()`

Unknown `kind` values should be rejected during schema parsing.

### `all`

`all` is a non-empty array of child schemas.

The node is valid only if **all** child schemas validate successfully.

Example:

```json
{
  "all": [
    { "kind": "string" },
    { "kind": "choice", "choices": ["a", "b", "c"] }
  ]
}
```

```{r}
schema <- schema_read('{
  "all": [
    { "kind": "string" },
    { "kind": "choice", "choices": ["a", "b", "c"] }
  ]
}')
good <- "a"
bad <- "z"

schema_validate(schema, good, mode = "test", name = "payload")
schema_validate(schema, bad, mode = "check", name = "payload")
```

### `any`

`any` is a non-empty array of child schemas.

The node is valid if **at least one** child schema validates successfully.

Example:

```json
{
  "any": [
    { "kind": "int", "lower": 0 },
    { "kind": "string", "pattern": "^[0-9]+$" }
  ]
}
```

```{r}
schema <- schema_read('{
  "any": [
    { "kind": "int", "lower": 0 },
    { "kind": "string", "pattern": "^[0-9]+$" }
  ]
}')
good <- "123"
bad <- TRUE

schema_validate(schema, good, mode = "test", name = "payload")
schema_validate(schema, bad, mode = "check", name = "payload")
```

### `one`

`one` is a non-empty array of child schemas.

The node is valid if **exactly one** child schema validates successfully.

Example:

```json
{
  "one": [
    { "kind": "string" },
    { "kind": "string", "pattern": "^[0-9]+$" }
  ]
}
```

```{r}
schema <- schema_read('{
  "one": [
    { "kind": "string" },
    { "kind": "string", "pattern": "^[0-9]+$" }
  ]
}')
good <- "abc"
bad <- TRUE

schema_validate(schema, good, mode = "test", name = "payload")
schema_validate(schema, bad, mode = "check", name = "payload")

bad <- "123"
schema_validate(schema, bad, mode = "check", name = "payload")
```

### `not`

`not` is a single child schema.

The node is valid only if the child schema does **not** validate successfully.

Example:

```json
{
  "not": { "kind": "null" }
}
```

```{r}
schema <- schema_read('{
  "not": { "kind": "null" }
}')
good <- "x"
bad <- NULL

schema_validate(schema, good, mode = "test", name = "payload")
schema_validate(schema, bad, mode = "check", name = "payload")
```

### `$ref`

`$ref` is a local definition reference.

Rules:

- `$ref` must be a string.
- Only local references of the form `#/$defs/name` are supported.
- A `$ref` node may optionally include `description`.
- Other node-level schema keys must not appear alongside `$ref`.

A `$ref` node looks like:

```json
{
  "$ref": "#/$defs/scalar_text",
  "description": "user-facing label"
}
```

## Shorthand check syntax inside combinators

Inside the child positions of:

- `all`
- `any`
- `one`
- `not`

the DSL allows a shorthand form for a `check` node:

```json
{ "kind": "string", "pattern": "^[0-9]+$" }
```

This is equivalent to:

```json
{
  "check": {
    "kind": "string",
    "pattern": "^[0-9]+$"
  }
}
```

Rules for shorthand check objects:

- They must contain `kind`.
- They may use any supported `check.kind`, including container kinds such as
  `list`, `data_frame`, `data_table`, and `tibble`.
- They must not contain any explicit primary operator keys:
  - `check`
  - `all`
  - `any`
  - `one`
  - `not`
  - `$ref`
- They must not contain any node-level adjunct keys:
  - `fields`
  - `groups`
  - `keys`
  - `description`
  - `$defs`
- They are only valid inside combinator child positions.

If a combinator branch needs node-level metadata or adjunct schema features,
write it as a full node instead of shorthand.

Allowed examples:

```json
{ "kind": "string", "pattern": "^[0-9]+$" }
```

```json
{ "kind": "list" }
```

```json
{ "kind": "data_frame", "min.rows": 1 }
```

Disallowed examples:

```json
{ "kind": "list", "keys": { "type": "unique" } }
```

```json
{ "kind": "list", "fields": { "value": { "check": { "kind": "string" } } } }
```

```json
{ "kind": "string", "description": "metadata is not allowed in shorthand" }
```

```json
{ "kind": "string", "$ref": "#/$defs/text" }
```

Example:

```json
{
  "any": [
    { "kind": "string" },
    {
      "check": { "kind": "list" },
      "fields": {
        "value": { "check": { "kind": "string" } }
      }
    }
  ]
}
```

```{r}
schema <- schema_read('{
  "any": [
    { "kind": "string" },
    {
      "check": { "kind": "list" },
      "fields": {
        "value": { "check": { "kind": "string" } }
      }
    }
  ]
}')
good <- list(value = "x")
bad <- list(value = 1L)

schema_validate(schema, good, mode = "test", name = "payload")
schema_validate(schema, bad, mode = "check", name = "payload")
```

## `$defs` and `$ref`

The DSL supports a minimal, local-reference reuse mechanism inspired by JSON
Schema.

### `$defs`

- `$defs` is only allowed at the root schema document.
- `$defs` must be an object whose values are complete schema nodes.
- `$defs` entries do not participate in validation directly; they are only
  targets for `$ref`.

Example:

```json
{
  "$defs": {
    "scalar_text": {
      "check": { "kind": "string" }
    }
  },
  "check": { "kind": "list" },
  "fields": {
    "name": { "$ref": "#/$defs/scalar_text" }
  },
  "rest": { "$ref": "#/$defs/scalar_text" }
}
```

```{r}
schema <- schema_read('{
  "$defs": {
    "scalar_text": {
      "check": { "kind": "string" }
    }
  },
  "check": { "kind": "list" },
  "fields": {
    "name": { "$ref": "#/$defs/scalar_text" }
  },
  "rest": { "$ref": "#/$defs/scalar_text" }
}')
good <- list(name = "alice", label = "admin")
bad <- list(name = "alice", label = 1L)

schema_validate(schema, good, mode = "test", name = "payload")
schema_validate(schema, bad, mode = "check", name = "payload")
```

### `$ref`

- A `$ref` node is replaced by the referenced schema node during parsing.
- A `$ref` node may optionally include `description`, which overrides the target
  node's description.

## `fields`

`fields` is only valid on nodes where:

- the primary operator is `check`
- `check.kind` is a container kind

`fields` must be a JSON object whose values are complete schema nodes.
The field name `"*"` has no special meaning; it is treated like any other
literal field name. Use `rest` for otherwise unspecified fields.

Example:

```json
{
  "check": { "kind": "list" },
  "fields": {
    "field1": { "check": { "kind": "string" } },
    "field2": { "check": { "kind": "int" } }
  }
}
```

```{r}
schema <- schema_read('{
  "check": { "kind": "list" },
  "fields": {
    "field1": { "check": { "kind": "string" } },
    "field2": { "check": { "kind": "int" } }
  }
}')
good <- list(field1 = "x", field2 = 1L)
bad <- list(field1 = "x", field2 = "1")

schema_validate(schema, good, mode = "test", name = "payload")
schema_validate(schema, bad, mode = "check", name = "payload")
```

`fields` does not make fields required. A missing field is simply not validated;
use `keys.must.include` when a field must be present.

### Pattern fields

`patterns` maps regular expressions to schema nodes. Patterns are applied only
to fields that are not already covered by `fields` or `groups`.

Example:

```json
{
  "check": { "kind": "list" },
  "patterns": {
    "^meta_": { "check": { "kind": "string" } },
    "_count$": { "check": { "kind": "int" } }
  }
}
```

```{r}
schema <- schema_read('{
  "check": { "kind": "list" },
  "patterns": {
    "^meta_": { "check": { "kind": "string" } },
    "_count$": { "check": { "kind": "int" } }
  }
}')
good <- list(meta_title = "report", row_count = 3L)
bad <- list(meta_title = 1L, row_count = 3L)

schema_validate(schema, good, mode = "test", name = "payload")
schema_validate(schema, bad, mode = "check", name = "payload")
```

If a field matches multiple patterns, it must satisfy every matching schema.

### Rest schema

`rest` is the child schema for otherwise unspecified fields.

Example:

```json
{
  "check": { "kind": "list" },
  "rest": { "check": { "kind": "string" } }
}
```

```{r}
schema <- schema_read('{
  "check": { "kind": "list" },
  "rest": { "check": { "kind": "string" } }
}')
good <- list(a = "x", b = "y")
bad <- list(a = "x", b = 1L)

schema_validate(schema, good, mode = "test", name = "payload")
schema_validate(schema, bad, mode = "check", name = "payload")
```

This means that any field not covered by `fields`, `groups`, or `patterns` must
satisfy the `rest` schema.

### Closed field sets with `keys`

`fields`, `groups`, and `patterns` validate matching fields, but they do not
close the set of allowed names by themselves. Use `keys.subset.of` to reject
unknown names, and `keys.must.include` when fields are required.

```json
{
  "check": { "kind": "list" },
  "keys": {
    "type": "unique",
    "must.include": ["field1", "field2"],
    "subset.of": ["field1", "field2"]
  },
  "fields": {
    "field1": { "check": { "kind": "string" } },
    "field2": { "check": { "kind": "int" } }
  }
}
```

```{r}
schema <- schema_read('{
  "check": { "kind": "list" },
  "keys": {
    "type": "unique",
    "must.include": ["field1", "field2"],
    "subset.of": ["field1", "field2"]
  },
  "fields": {
    "field1": { "check": { "kind": "string" } },
    "field2": { "check": { "kind": "int" } }
  }
}')
good <- list(field1 = "x", field2 = 1L)
bad <- list(field1 = "x", field2 = 1L, extra = "y")

schema_validate(schema, good, mode = "test", name = "payload")
schema_validate(schema, bad, mode = "check", name = "payload")
```

### Precedence rule

Container child schemas are applied in this order:

1. exact `fields` and `groups`
2. `patterns` for fields not covered by exact rules
3. `rest` for fields not covered by exact or pattern rules

## `positions`

`positions` is only valid on unnamed container nodes:

- the primary operator is `check`
- `check.kind` is a container kind
- `keys.type` is `"unnamed"`

`positions` is an unnamed array of schema nodes. It follows JSON Schema
`prefixItems` semantics: a declared position is validated only when the input
actually has that position. Missing positions do not fail by themselves; use
`len`, `min.len`, or `max.len` inside the `check` rule when length matters.

Example, similar to a CrossRef `date-parts` item:

```json
{
  "check": { "kind": "list", "min.len": 1, "max.len": 3 },
  "keys": { "type": "unnamed" },
  "positions": [
    { "check": { "kind": "int", "lower": 0 } },
    { "check": { "kind": "int", "lower": 1, "upper": 12 } },
    { "check": { "kind": "int", "lower": 1, "upper": 31 } }
  ]
}
```

```{r}
schema <- schema_read('{
  "check": { "kind": "list", "min.len": 1, "max.len": 3 },
  "keys": { "type": "unnamed" },
  "positions": [
    { "check": { "kind": "int", "lower": 0 } },
    { "check": { "kind": "int", "lower": 1, "upper": 12 } },
    { "check": { "kind": "int", "lower": 1, "upper": 31 } }
  ]
}')
good <- list(2024L, 6L, 14L)
bad <- list(2024L, 13L, 14L)

schema_validate(schema, good, mode = "test", name = "payload")
schema_validate(schema, bad, mode = "check", name = "payload")
```

When `positions` is combined with `rest`, declared positions are validated first
and all remaining positions are validated with `rest`:

```json
{
  "check": { "kind": "list" },
  "keys": { "type": "unnamed" },
  "positions": [
    { "check": { "kind": "string" } },
    { "check": { "kind": "int" } }
  ],
  "rest": { "check": { "kind": "number" } }
}
```

```{r}
schema <- schema_read('{
  "check": { "kind": "list" },
  "keys": { "type": "unnamed" },
  "positions": [
    { "check": { "kind": "string" } },
    { "check": { "kind": "int" } }
  ],
  "rest": { "check": { "kind": "number" } }
}')
good <- list("run", 1L, 2.5, 3)
bad <- list("run", 1L, "extra")

schema_validate(schema, good, mode = "test", name = "payload")
schema_validate(schema, bad, mode = "check", name = "payload")
```

With `keys.type = "unnamed"`, `positions` and `rest` are allowed. `fields`,
`groups`, and `patterns` are named-object constraints and must not be mixed with
unnamed array semantics.

## `groups`

`groups` is an optional array available only on nodes where:

- the primary operator is `check`
- `check.kind` is a container kind

Each group item must be a named object with:

- `names`: a non-empty character vector of field names
- a complete schema node written directly beside `names`

`groups` is expanded during parsing into ordinary named fields that behave exactly
like entries written explicitly in `fields`.

Rules:

- group names must be unique within a group
- group names must not overlap across groups
- group names must not overlap with explicit `fields`
- each group item must contain exactly one primary operator in addition to `names`
- a group item must not wrap its target node inside an anonymous nested list
- a group item must not contain both `$ref` and `check` at the same level
- if present, `description` must be a non-empty string

If a group item provides `description`, it is copied to every expanded field in
that group and overrides the referenced node's own description.

Valid examples:

```json
{
  "names": ["name", "label"],
  "$ref": "#/$defs/text"
}
```

```json
{
  "names": ["name", "label"],
  "all": [
    { "$ref": "#/$defs/text" },
    { "check": { "kind": "string" } }
  ]
}
```

Groups can also share a container schema. This is useful after compacting API
payload schemas where several fields have the same nested structure:

```json
{
  "names": ["issued", "created", "published-print"],
  "check": { "kind": "list" },
  "fields": {
    "date-parts": {
      "check": { "kind": "list" },
      "keys": { "type": "unnamed" },
      "rest": {
        "check": { "kind": "list" },
        "keys": { "type": "unnamed" },
        "rest": { "check": { "kind": "int" } }
      }
    }
  }
}
```

Invalid example (R-list shape):

```r
list(
  names = c("name", "label"),
  list(`$ref` = "#/$defs/text", check = list(kind = "string"))
)
```

The invalid form above is rejected because:

- the target node is wrapped in an anonymous nested list
- the inner object contains more than one primary operator

Example:

```json
{
  "$defs": {
    "flag_param": {
      "check": { "kind": "flag", "null.ok": true }
    }
  },
  "check": { "kind": "list" },
  "groups": [
    {
      "names": ["latest", "distrib", "replica"],
      "$ref": "#/$defs/flag_param",
      "description": "boolean control parameters"
    }
  ],
  "rest": { "check": { "kind": "string" } }
}
```

```{r}
schema <- schema_read('{
  "$defs": {
    "flag_param": {
      "check": { "kind": "flag", "null.ok": true }
    }
  },
  "check": { "kind": "list" },
  "groups": [
    {
      "names": ["latest", "distrib", "replica"],
      "$ref": "#/$defs/flag_param",
      "description": "boolean control parameters"
    }
  ],
  "rest": { "check": { "kind": "string" } }
}')
good <- list(latest = TRUE, distrib = FALSE, replica = NULL, project = "CMIP6")
bad <- list(latest = "yes", distrib = FALSE, replica = NULL, project = "CMIP6")

schema_validate(schema, good, mode = "test", name = "payload")
schema_validate(schema, bad, mode = "check", name = "payload")
```

## `keys`

`keys` is only valid on container `check` nodes.

`keys` encodes arguments for:

```r
checkmate::check_names(names(x), ...)
```

### Minimal object form

Even when only the name-check type is needed, `keys` is written as an object:

```json
{ "type": "..." }
```

The `type` value is passed to the `type` argument of
`checkmate::check_names(names(x), ...)`.

Example:

```json
{
  "check": { "kind": "list" },
  "keys": { "type": "unique" }
}
```

```{r}
schema <- schema_read('{
  "check": { "kind": "list" },
  "keys": { "type": "unique" }
}')
good <- list(a = 1, b = 2)
bad <- list(a = 1, a = 2)

schema_validate(schema, good, mode = "test", name = "payload")
schema_validate(schema, bad, mode = "check", name = "payload")
```

### Object form

If `keys` is an object:

- all keys are passed directly to `checkmate::check_names(names(x), ...)`
- `keys.type` is the only way to specify the name-check type

Example:

```json
{
  "check": { "kind": "list" },
  "keys": {
    "type": "unique",
    "must.include": ["kind", "value"],
    "subset.of": ["kind", "value", "negate"]
  }
}
```

This represents:

```r
checkmate::check_names(
  names(x),
  type = "unique",
  must.include = c("kind", "value"),
  subset.of = c("kind", "value", "negate")
)
```

```{r}
schema <- schema_read('{
  "check": { "kind": "list" },
  "keys": {
    "type": "unique",
    "must.include": ["kind", "value"],
    "subset.of": ["kind", "value", "negate"]
  }
}')
good <- list(kind = "facet", value = "tas", negate = FALSE)
bad <- list(kind = "facet", value = "tas", extra = TRUE)

schema_validate(schema, good, mode = "test", name = "payload")
schema_validate(schema, bad, mode = "check", name = "payload")
```

## `description`

`description` is metadata only.

It does not affect validation and may be used for documentation, debugging, or
pretty-printing.

`description` may appear on any full schema node and on `$ref` nodes.

For canonical serialization via `as.list()`, `description` is emitted first
inside ordinary schema nodes. At the top-level document, `version` remains first
when present, then root `description`, then `$defs`, then the root operator keys.

## Validation order for one node

Validation depends on the node's primary operator.

### `check` node

Validation of a `check` node proceeds in this order:

1. run the primary `assert_<kind>()`
2. apply optional `keys` handling
3. if `check.kind` is a named container, recursively validate exact fields,
   pattern fields, and rest fields
4. if `check.kind` is an unnamed container, recursively validate positions, then
   rest positions

### `all`

Validate each child schema in order.

- If any child fails, the `all` node fails immediately.
- If all children pass, the `all` node passes.

### `any`

Validate child schemas until one succeeds.

- If a child succeeds, the `any` node passes immediately.
- If all children fail, the `any` node fails.

### `one`

Validate child schemas while counting successful branches.

- If exactly one child succeeds, the `one` node passes.
- If zero children succeed, the `one` node fails.
- If more than one child succeeds, the `one` node fails.

### `not`

Validate the child schema and invert the result.

- If the child passes, the `not` node fails.
- If the child fails, the `not` node passes.

### `$ref`

Resolve the reference and validate against the target schema.

## Formal validation rules

When parsing the JSON DSL itself, the following should be enforced:

1. Every node must contain exactly one primary operator from `check`, `all`,
   `any`, `one`, `not`, `$ref`.
2. `check` must be an object when present.
3. `check.kind` must be a non-empty string.
4. `check.kind` must name a supported `checkmate::assert_*()` suffix.
5. `all`, `any`, and `one` must be non-empty arrays when present.
6. `not` must be a single schema node.
7. Child entries of `all`, `any`, `one`, and `not` must be either:
   - a complete schema node, or
   - a shorthand check object.
8. A shorthand check object must contain `kind`.
9. A shorthand check object may use any supported `check.kind`.
10. A shorthand check object must not contain explicit primary operator keys.
11. A shorthand check object must not contain node-level adjunct keys.
12. `fields` is only allowed when the node is a `check` node and
    `check.kind` is a container kind.
13. `groups` is only allowed when the node is a `check` node and
    `check.kind` is a container kind.
14. `fields` must be an object when present.
15. `groups` must be an array when present.
16. `patterns` must be an object when present.
17. `positions` must be an unnamed array when present.
18. `positions` requires `keys.type = "unnamed"`.
19. If `keys.type = "unnamed"`, `fields`, `groups`, and `patterns` must not be
    present.
20. `keys` is only allowed on container `check` nodes.
21. `keys` must be an object.
22. All `keys` arguments are forwarded to
    `checkmate::check_names()`.
23. `keys.check` is invalid.
24. Each `groups[]` item must contain `names` plus a complete schema node.
25. The schema node inside `groups[]` follows the same primary-operator and
    keyword rules as any other schema node.
26. `groups[]` items must not wrap the target node inside an anonymous nested
    list.
27. `version` is only allowed at the root schema document.
28. `version` must be a non-empty string when present.
29. `$defs` is only allowed at the root schema document.
30. `$defs` must be an object whose values are complete schema nodes.
31. `$ref` must be a local string reference of the form `#/$defs/name`.
32. A `$ref` target must exist in the current document's `$defs`.
33. A `$ref` node must not contain other schema keys except `description`.
34. `description` must be a non-empty string when present.

## Additional examples

### Parameter-like schema

```json
{
  "check": { "kind": "list" },
  "keys": {
    "type": "unique",
    "subset.of": ["project", "activity_id", "experiment_id"]
  },
  "rest": {
    "check": { "kind": "list" },
    "keys": {
      "type": "unique",
      "must.include": ["kind", "value"],
      "subset.of": ["kind", "value", "negate"]
    },
    "fields": {
      "kind": {
        "check": {
          "kind": "choice",
          "choices": ["facet", "datetime_start", "datetime_stop"]
        }
      },
      "value": {
        "check": {
          "kind": "atomic",
          "any.missing": false
        }
      },
      "negate": {
        "check": {
          "kind": "flag",
          "null.ok": true
        }
      }
    }
  }
}
```

```{r}
param_schema <- schema_read('{
  "check": { "kind": "list" },
  "keys": {
    "type": "unique",
    "subset.of": ["project", "activity_id", "experiment_id"]
  },
  "rest": {
    "check": { "kind": "list" },
    "keys": {
      "type": "unique",
      "must.include": ["kind", "value"],
      "subset.of": ["kind", "value", "negate"]
    },
    "fields": {
      "kind": {
        "check": {
          "kind": "choice",
          "choices": ["facet", "datetime_start", "datetime_stop"]
        }
      },
      "value": {
        "check": {
          "kind": "atomic",
          "any.missing": false
        }
      },
      "negate": {
        "check": {
          "kind": "flag",
          "null.ok": true
        }
      }
    }
  }
}')
good <- list(
    project = list(kind = "facet", value = "CMIP6", negate = FALSE),
    activity_id = list(kind = "facet", value = "ScenarioMIP")
)
bad <- list(
    project = list(kind = "unknown", value = "CMIP6"),
    activity_id = list(kind = "facet", value = "ScenarioMIP")
)

schema_validate(param_schema, good, mode = "test", name = "payload")
schema_validate(param_schema, bad, mode = "check", name = "payload")
```

## Common mistakes

- Do not mix primary operators in one node. A node cannot contain both `check`
  and `all`, for example.
- Do not put field names directly beside `check`. Field names belong under
  `fields` or `groups`.
- Do not use JSON Schema keywords such as `type`, `properties`, `required`, or
  `additionalProperties` unless they are checkmate arguments inside `check` or
  `keys`.
- Do not expect `fields` to imply required fields. Add a `keys` rule such as
  `"must.include": ["id", "name"]` when fields must be present.
- Do not use `$defs` inside nested nodes. Definitions live at the document root.
