---
title: "Introduction to schemate"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to schemate}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

`schemate` is built around one lifecycle:

1. infer a conservative schema from an example object;
2. compact repeated observed structure;
3. edit the schema until it expresses the real input contract;
4. write the schema to JSON;
5. read the schema back where it is needed;
6. validate new input.

## Nested List Contract

The most useful `schemate` workflows start with a nested R object such as a
package configuration, a model payload, or a JSON-like API response.

```{r}
library(schemate)

payload <- list(
    request = list(id = "run-001", retry = FALSE),
    items = list(
        list(id = 1L, label = "alpha", tags = list("r", "schema")),
        list(id = 2L, label = "beta", tags = list("validation"))
    )
)

schema <- payload |>
    schema_infer(keys = "named", arrays = "rest") |>
    schema_compact()

schema
```

`arrays = "rest"` treats unnamed lists as homogeneous arrays. The observed item
schemas are stored in `rest`, and `schema_compact()` merges compatible observed
alternatives into one maintainable item schema.

## Refine

Inference captures observed structure. It does not guess business rules, so the
next step is to refine the parts of the contract that matter.

```{r}
schema <- schema |>
    schema_set_desc("$", "Example request payload.") |>
    schema_replace("$request$id", schema_check("string", min.chars = 1)) |>
    schema_replace("$items$rest$id", schema_check("int", lower = 1)) |>
    schema_set_rest(schema_check("string", min.chars = 1), path = "$items$rest$tags")

schema
```

Paths use `$` for the root node. Bare field paths such as `$request$id` traverse
container fields. Inferred unnamed array schemas are reached through `rest`, as
in `$items$rest$id`.

## Write And Read

`schema_read()` and `schema_write()` require the suggested package `jsonlite`.

```{r}
path <- tempfile(fileext = ".json")
schema_write(schema, path)

restored <- schema_read(path)
restored
```

Below is an example schema file shipped with this package.
`system.file()`.

```{r}
person_schema <- system.file("extdata", "person-schema.json", package = "schemate")
schema_read(person_schema)
```

## Validate

```{r}
good <- payload
restored |>
    schema_validate(good, mode = "test")

bad <- payload
bad$items[[1L]]$id <- 0L
restored |>
    schema_validate(bad, mode = "check", name = "payload")
```

Diagnostics include a path prefix. A message starting with `payload$items[[1]]$id`
means the root object named `payload` failed inside the first item at field `id`.
Messages from leaf checks come from checkmate, while container messages are
produced by `schemate` when fields, names, branches, or references do not match.

## Data Frame Inputs

Data frames are also container objects. `schemate` is not a data quality
reporting framework; use it when you want an input schema that can be inferred,
edited, saved, and reused.

```{r}
scores <- data.frame(
    id = 1:3,
    name = c("alice", "bob", "carol"),
    score = c(9.5, 8.0, 7.5)
)

score_schema <- scores |>
    schema_infer(keys = "required") |>
    schema_replace("$id", schema_check("integerish", any.missing = FALSE)) |>
    schema_replace("$score", schema_check("numeric", lower = 0, upper = 10))
score_schema

score_schema |>
    schema_validate(scores, mode = "test")

bad_scores <- transform(scores, score = as.character(score))
score_schema |>
    schema_validate(bad_scores, mode = "check", name = "scores")
```

## Validation Modes

Use validation modes according to the caller:

| Mode | Use when |
| --- | --- |
| `assert` | invalid input should stop execution |
| `check` | you want a diagnostic string |
| `test` | control flow needs a boolean |
| `expect` | tests should receive a testthat-style expectation object |
